---
title: "Using genetrackR with airway RNA-seq data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using genetrackR with airway RNA-seq data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(genetrackR)
library(airway)
library(GenomicFeatures)
library(SummarizedExperiment)
library(ggplot2)
library(R.utils)
```

## Introduction

This vignette demonstrates how to use the `genetrackR` package for gene-level RNA-seq analysis using real data from the `airway` package. We walk through the full analysis pipeline: counting reads, normalization, visualization, and exporting results.

## 1. Prepare input data

We use a BAM file included in the `airway` package and a gene annotation GTF file that comes with this package.

```{r}
bam_file <- system.file("extdata", "SRR1039508_subset.bam", package = "airway")

gtf_gz <- system.file("extdata", "Homo_sapiens.GRCh38.114.gtf.gz", package = "genetrackR")
gtf_file <- file.path(tempdir(), "annotation.gtf")
if (!file.exists(gtf_file)) {
  R.utils::gunzip(gtf_gz, destname = gtf_file, remove = FALSE)
}
```

## 2. Create TxDb from GTF annotation

```{r}
txdb <- GenomicFeatures::makeTxDbFromGFF(gtf_file, format = "gtf")
```

## 3. Count reads per gene

```{r}
se <- countReads(bam_files = bam_file, txdb = txdb)
se
```

## 4. Normalize counts to TPM

```{r}
raw_counts <- assay(se, "counts")
tpm <- normalizeCounts(raw_counts, txdb = txdb, method = "TPM")
head(tpm)
```

## 5. Visualize gene expression

We show gene expression of the first gene across all samples, and the top 20 expressed genes in the first sample.

```{r}
p1 <- plotGeneExpression(tpm, geneID = rownames(tpm)[1])
p2 <- plotGeneExpression(tpm, sampleID = colnames(tpm)[1])
```

## 6. Export results

```{r}
exportResults(tpm, filename = "counts_TPM.csv", path = tempdir(), type = "table")
exportResults(p1, filename = "gene_plot.png", path = tempdir(), type = "plot")
```

The resulting files can be found in your temporary directory.

## Conclusion

With `genetrackR`, you can analyze RNA-seq BAM files, normalize gene counts, generate expression plots, and export results — all in a streamlined, reproducible workflow using standard Bioconductor tools.
